---
title: "Alcohol consumption by Portuguese students"
output: html_notebook
author: Shiva Kumar Pendem, Sricharan Cheeti, Kaushik Parvathaneni
---
```{r}
library(dplyr)
library(tidyverse)
library(tidyr)
library(corrplot)
library(plotly)
library(caTools)
library(car)
```


```{r}
Mat = read.csv('studentMat.csv')

```

### Preprocessing
#### Converting Non numeric to Numeric

```{r }
non_numeric_columns <- names(Mat)[sapply(Mat, function(col) !is.numeric(col))]
cat("These are non-numeric columns:", non_numeric_columns, "\n\n")

numeric_columns <- names(Mat)[sapply(Mat, is.numeric)]
cat("These are numeric columns:", numeric_columns, "\n")


```
We have to convert Non numeric data to numeric data to find the correlation coeff

```{r}
# Mat$sex <- ifelse(Mat$sex == 'M', 1, 0)
```

```{r}
# non_numeric_columns <- c('school', 'address', 'famsize', 'Pstatus', 'Mjob', 'Fjob', 'reason', 'guardian', 'schoolsup', 'famsup', 'paid', 'activities', 'nursery', 'higher', 'internet', 'romantic')

# for(col in non_numeric_columns) {
#   if(length(unique(Mat[[col]])) == 2) {
#     # Label Encoding
#     levels <- unique(Mat[[col]])
#     Mat[[col]] <- ifelse(Mat[[col]] == levels[1], 1, 0)
#   } else {
#     # One-Hot Encoding
#     Mat <- Mat %>%
#       mutate(!!col := as.factor(!!sym(col))) %>%
#       spread(!!col, !!col, fill = 0, convert = TRUE)
#   }
# }
non_numeric <- names(Mat)[sapply(Mat, function(col) !is.numeric(col))]

for (col in non_numeric){
  print(paste("The column ", col, "has ", length(unique(Mat[[col]])), ' unique values they are: '))
  print((unique(Mat[[col]])))
  cat('\n')
}
```

```{r Arranging Binary and non binary columns}
binary_columns <- c()
multi_unique_columns <- c()

# Loop through each column and categorize based on the number of unique values
for (col in non_numeric) {
    num_unique <- length(unique(Mat[[col]]))
    
    if (num_unique == 2) {
        binary_columns <- c(binary_columns, col)
    } else if (num_unique > 2) {
        multi_unique_columns <- c(multi_unique_columns, col)
    }
}
# for (col in binary_columns) {
#   unique_vals <- unique(Mat[[col]])
#   Mat[[col]] <- ifelse(Mat[[col]] == unique_vals[1], 0, 1)
# }
print(paste("Binary columns:", paste(binary_columns, collapse = ", ")))
cat('\n')
print(paste("Multi unique value columns:", paste(multi_unique_columns, collapse = ", ")))
```


```{r Turing binary columns numeric}

for (col in binary_columns) {
  if ("yes" %in% Mat[[col]] && "no" %in% Mat[[col]]) {
    # If the column has 'yes' and 'no' values
    Mat[[col]] <- ifelse(Mat[[col]] == "yes", 1, 0)
  } else {
    # For other binary columns
    unique_vals <- unique(Mat[[col]])
    Mat[[col]] <- ifelse(Mat[[col]] == unique_vals[1], 0, 1)
    print(paste(unique_vals[1],':0', unique_vals[2],':1'))
  }
}

```
```{r Turing multivalued columns numeric}
# Perform one-hot encoding for each column in multi_unique_columns
for (col in multi_unique_columns) {
  # Create a one-hot encoded matrix for the column
  formula_str <- paste("~ 0 +", col)
  one_hot <- model.matrix(as.formula(formula_str), data = Mat)
  
  # Convert matrix to data frame and set column names
  one_hot_df <- as.data.frame(one_hot)
  colnames(one_hot_df) <- gsub("^.\\.", col, colnames(one_hot_df))
  
  # Bind the new one-hot encoded columns to the original data frame
  Mat <- cbind(Mat, one_hot_df)
  
  # Remove the original column
  Mat[[col]] <- NULL
}

# View the first few rows of the processed data to verify
head(Mat)

```
```{r}
cor_matrix <- cor(Mat, use = "complete.obs")
```
```{r}
corrplot(cor_matrix, method = "circle")
```
```{r}
plot_ly(
  x = colnames(cor_matrix),
  y = rownames(cor_matrix),
  z = cor_matrix,
  type = "heatmap",
  colorscale = "Viridis"
) %>%
  layout(title_text = "Correlation Matrix Heatmap")

```
```{r}
G3_correlation <- cor_matrix[,'G3']
library(plotly)

plot_ly(x = names(G3_correlation), y = G3_correlation, type = 'bar') %>%
  layout(title = "Correlation of G3 with Other Variables", yaxis = list(title = "Correlation Coefficient"))

```
```{r}
print(G3_correlation)
```
```{r}
# Sorting correlations in decreasing order to get top positive correlations
top_positive <- sort(G3_correlation, decreasing = TRUE)[1:10]

# Sorting correlations in increasing order to get top negative correlations
top_negative <- sort(G3_correlation)[1:10]

print("Top Positive Correlations:")
print(top_positive)

print("Top Negative Correlations:")
print(top_negative)

```
```{r}
# Sort G3_correlation in increasing order
sorted_names <- names(sort(G3_correlation))

# Convert variable names to a factor with levels specified by the sorted order
factor_names <- factor(names(G3_correlation), levels = sorted_names)

# Plot using plotly
plot_ly(x = factor_names, y = G3_correlation, type = 'bar') %>%
  layout(title = "Correlation of G3 with Other Variables in Increasing Order", 
         yaxis = list(title = "Correlation Coefficient"))


```

```{r Alc_consumtion}
Mat_alc <- Mat

```

```{r}
Mat_alc$Avg_alc <- (Mat_alc$Dalc + Mat_alc$Walc)/2
```

```{r}
Mat_alc[,c("Avg_alc", "Dalc", "Walc")]
```
```{r}
Mat_alc <- Mat_alc %>% select(-Dalc, -Walc)
```
```{r}
Mat_alc
```
```{r}
Mat_alc <- Mat_alc %>% select(-G1, -G2,-G3)
```

```{r}
corr_mat = cor(Mat_alc)
corrplot(corr_mat)
```

```{r}
Avg_alc_correlation <- corr_mat[,'Avg_alc']


plot_ly(x = names(Avg_alc_correlation), y = Avg_alc_correlation, type = 'bar') %>%
  layout(title = "Correlation of Avg_alc with Other Variables", yaxis = list(title = "Correlation Coefficient"))
```

```{r}

sorted_names <- names(sort(Avg_alc_correlation))


factor_names <- factor(names(Avg_alc_correlation), levels = sorted_names)

plot_ly(x = factor_names, y = Avg_alc_correlation, type = 'bar') %>%
  layout(title = "Correlation of Avg_alc with Other Variables in Increasing Order", 
         yaxis = list(title = "Correlation Coefficient"))

```

```{r}
Avg_alc_correlation
```

```{r}
Mat_alc
```

```{r}
# Assuming your data is in a variable named 'Mat_alc'



# 2. Split the data into training and testing sets
set.seed(123) # Setting seed for reproducibility
split = sample.split(Mat_alc$Avg_alc, SplitRatio = 0.8)
train_data = subset(Mat_alc, split == TRUE)
test_data = subset(Mat_alc, split == FALSE)

# 3. Train a linear regression model using the training set
model <- lm(Avg_alc ~ ., data = train_data)  # The dot means we are using all other columns as predictors

# 4. Evaluate the model using the testing set
predictions = predict(model, newdata = test_data)
mse = mean((predictions - test_data$Avg_alc)^2)  # Mean Squared Error

print(mse)

# Additionally, you can print a summary of the model to inspect coefficients and other statistics
print(summary(model))

```



```{r}
# Removing redudant variables for 1-hot encoding 
Mat_alc <- Mat_alc %>% select(-guardianother, -reasonreputation, -Fjobteacher, -Mjobteacher)
```


```{r}
set.seed(123) 
split = sample.split(Mat_alc$Avg_alc, SplitRatio = 0.8)
train_data = subset(Mat_alc, split == TRUE)
test_data = subset(Mat_alc, split == FALSE)


model <- lm(Avg_alc ~ ., data = train_data)   

predictions = predict(model, newdata = test_data)
mse = mean((predictions - test_data$Avg_alc)^2)  

print(mse)


print(summary(model))
```
```{r}

vif_model <- vif(model)  
print(vif_model)
```
```{r}
vif_model[vif_model >= 5]
```
Dropping Fjobother because the VIF value > 5
Generally, 
* 1 = not correlated.
* Between 1 and 5 = moderately correlated.
* Greater than 5 = highly correlated.

```{r}
vif_model[vif_model > 3 & vif_model < 5]
```
```{r}
vif_model[vif_model <=1]
```



```{r}
Mat_alc <- Mat_alc %>% select(-Fjobother)
```

```{r}
set.seed(123) 
split = sample.split(Mat_alc$Avg_alc, SplitRatio = 0.8)
train_data = subset(Mat_alc, split == TRUE)
test_data = subset(Mat_alc, split == FALSE)


model <- lm(Avg_alc ~ ., data = train_data)   

predictions = predict(model, newdata = test_data)
mse = mean((predictions - test_data$Avg_alc)^2)  

print(mse)


print(summary(model))
```

Important Features based on the Significance codes
sex, address, traveltime, paid, activities, nursery, famrel, goout, health, Fjobservices

```{r}
model_summary <- summary(model)
estimates <- model_summary$coefficients[, "Estimate"]
ordered_estimates_desc <- estimates[order(-estimates)]
print(ordered_estimates_desc)

```


```{r}
library(ggplot2)

# Convert ordered estimates to a data frame
df_estimates <- data.frame(Predictor = names(ordered_estimates_desc),
                           Estimate = ordered_estimates_desc)

# Plot using ggplot2
plot <- ggplot(df_estimates, aes(x = reorder(Predictor, Estimate), y = Estimate)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  coord_flip() + 
  labs(title = "Ordered Estimates from the Model",
       x = "Predictors",
       y = "Coefficient Value") +
  theme_minimal()

plot

```

